---
title: "Do people who frequently listen to certain music genres report better or worse mental health outcomes??"
subtitle: "Assignment 4 ETC5512"
author: "Yilin Zhang"
format: html
editor: visual
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data and Documenting

### Research Question

Can music help improve mental health, and if so, do different music genres have different impacts on specific mental health issues?

-   Whether frequent listening to certain genres if associated with lower anxiety, depression, insomnia, or OCD levels.

-   Whether certain genres are more effective for specific mental health concerns.

-   Whether listening conditions, such as listening while working or total hours per day might affect the mental health outcomes of music.

These questions aim to provide insights into how music could be a personalized, feasible tool to improve emotional well-being.

### Data Set and Suitability

In order to solve my doubts, I uses the "Music & Mental Health Survey" data set, which is published on Kaggle by Catherine Rasgaitis. This data set contains 736 public questionnaire surveys from Google.

This data set is highly suitable as the foundation for conducting research.

-   Listening preference: frequency of listening to specific genres, Music style preference, Primary streaming platform.

-   Mental health indicators: self-reported anxiety, depression, insomnia, and OCD.

-   Listening habits: whether respondent listen to music while work, play an instrument regularly, how many hours per day they listen and so on.

These data is helpful for exploring the positive impact of different genres of music on mental health.

### Data Description

-   Data set name: Music & Mental Health Survey

-   Source: \[Kaggle Dataset\](<https://www.kaggle.com/datasets/catherinerasgaitis/mxmh-survey-results>)

-   

### Data Preparation

```{r}
library(tidyverse)
library(janitor)

# reading data
music_raw <- read_csv("data/mxmh_survey_results.csv") |> 
  clean_names()

# transfrom frequency text into numerical value
freq_map <- c("Never" = 0, "Rarely" = 1, "Sometimes" = 2, "Very frequently" = 3)

# create and name new columns
music <- music_raw |> 
  mutate(across(starts_with("frequency_"), ~ freq_map[.x], .names = "num_{.col}"))

# only keep the records which are not missing
music_clean <- music |> 
  filter(!is.na(anxiety), !is.na(depression), !is.na(insomnia), !is.na(ocd))

glimpse(music_clean)

```

### Does listening to certain gernes help reduce anxiety?
```{r}
music_long <- music_clean |> 
  select(anxiety, starts_with("num_frequency_")) |> 
  pivot_longer(cols = starts_with("num_frequency_"),
               names_to = "genre",
               values_to = "frequency") |> 
  mutate(
    genre = str_remove(genre, "num_frequency_"),
    genre = str_replace_all(genre, "_", " ") |> str_to_title()
  )

filtered <- music_long |> filter(frequency >= 2)

ggplot(filtered, aes(x = reorder(genre, anxiety, median), y = anxiety)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Anxiety levels among frequent listeners of different music genres",
    x = "Music Genre",
    y = "Anxiety Score (0–10)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
### Are some genres more helpful for other mental health issues?

### Remember

Please mention any additional files that you want the markers to review, eg. read me, meta data etc.

<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### What's in this section

Here is where you should write your blogpost! (Task 2)

### Blogpost Structure

There is no strict structure for you to follow. However, here is a skeleton structure you may find helpful.

1.  Title (is set at the top of this document)
2.  Motivation\
3.  Data\
4.  Analysis\
5.  Conclusions\
6.  References

<!-- Task 3 Tab: Behind the Scenes -  -->

## Behind the Scenes

### What's in this section

Here is where you should tell us about your reflection on your analysis (Task 3).

Again, these are the details about **your** perspective and the gritty details behind the scenes of your analysis.
:::
